# Week 2: Wrangle & Analyse

*This week we follow through the examples in Exercise Set 2 and learn about how to do linear regression and beyond with R.*

A few words about the data set:

-   Assuming that you the reader have no no previous knowledge of the data -- don't worry, neither had I just a few minutes ago -- from the appearance of it, there was a survey about study methods and "attitude" survey filled in by just under 200 Finnish pre-retirees. There were also exam points, so they were perhaps also given some kind of tests. More details [here](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS2-meta.txt).

-   Given the schema of the data, now reverse-guessing the possible hypotheses, it's possible that the claim was some attitude or strategy predicted exam scores. That's kind of reasonable, maybe people who believe stats can*"hyötyä"* would score higher than people who don't believe that.

```{r}
date()
```

### Step 1: Read the data

```{r}
library(readr)
data <- read_csv('data/learning_2014.csv')
str(data)
dim(data)
```

The data contains 166 survey responses. Each response contains 7 columns (variables), namely: gender, age of the participant, attitude score (being the average of 10 attitude questions about attitudes towards learning about statistics), deep/surface/strategic metrics (the average of different categories of questions related to learning approaches or methods) -- see more details in this [document](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-meta.txt) and the points from some exam or test.

So the general gist is -- you may have guessed it -- what influences study outcomes in statistics? Is it attitude towards learning? Is it approach and life hacks in learning? Let's find out.

### Step 2: Graphical overview

"Show a graphical overview", a "plot of everything":

```{r message = FALSE}
library(GGally)
library(ggplot2)
p <- ggpairs(data, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20)))
p
```

The histograms (on the left-hand side) are quite useful for showing the distributions, we can see for example at a glance there are actually a lot of younger students who answered the survey, even though the first few rows have older people.

The box plots show how gender correlate with the other variables.

The scatter plots are not so easy for humans to eyeball the trends.

But, what's most interesting for us may be the vertical strip of printouts where "points" correlate with other variables.

### Step 3: Fit a model

Let's pick the three largest values: `attitude`, `stra`, `surf`

```{r}
p1 <- ggplot(data, aes(x = attitude, y = points))
p2 <- p1 + geom_point()
p2
```

Seems a bit all over the place. Let's add a regression line:

```{r}
p3 <- p2 + geom_smooth(method = "lm")
# Draw plot
p3
```

Now, onto fitting a regression model with all three variables!

```{r}
multi_model <- lm(points ~ attitude + stra + surf, data = data)
multi_model
summary(multi_model)
```

### Step 4: Interpretation

Looking at the fitted coefficients (r-squareds), attitude has a big influence, stra has a much smaller influence, surf has a smaller still and negative influence.

### Step 5: Graphical validation

```{r}
# Residuals vs Fitted values (1)
p_a = plot(multi_model, which = 1)
```

```{r}
# Normal QQ-plot (2)
p_b = plot(multi_model, which = 2)
```

```{r}
# Residuals vs Leverage (5)
p_c = plot(multi_model, which = 5)
```

The assumption of the model is that attitude, stra, and surf influence points. Looking at the visual output, it's not conclusively clear-cut.
