# Week 5

*This week we follow through the examples in Exercise Set 5.*

```{r}
date()
```

------------------------------------------------------------------------

### Step 0

Create this file and include it in `index.Rmd`.

### Step 1

> "Move the country names to rownames"

> "Show a graphical overview of the data"

> "Show summaries of the variables in the data"

> "Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them"

### Step 2

> "Perform principal component analysis (PCA) on the raw (non-standardized) human data"

> "Show the variability captured by the principal components"

> "Draw a biplot displaying the observations by the first two principal components along with arrows representing the original variables"

### Step 3

> "Standardize the variables in the human data and repeat the above analysis"

> "Interpret the results of both analysis (with and without standardizing)" *"Are the results different? Why or why not?"*

> "Include captions (brief descriptions) in your plots where you describe the results by using not just your variable names, but the actual phenomena they relate to"

### Step 4

> "Give your personal interpretations of the first two principal component dimensions based on the biplot drawn after PCA on the standardized human data"

### Step 5

> "Load the tea dataset and convert its character variables to factors"

```{r}
tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)
```

> "Explore the data briefly: look at the structure and the dimensions of the data"

> "Use View(tea) to browse its contents, and visualize the data"

> "Use Multiple Correspondence Analysis (MCA) on the tea data (or a subset of columns)"

> "Interpret the results of the MCA"

> "Draw the variable biplot of the analysis"

> "Comment on the output of the plots"
